---
title: "Basic Benthic Template"
author: "Emma Jones"
date: "4/30/2019"
output: 
  html_document
  
---

```{r setup, include= FALSE, message= FALSE}
knitr::opts_chunk$set(echo = FALSE) # This line makes sure no R chunks are displayed in the output report

# Bring in necessary packages
library(tidyverse)
library(knitr)
library(reshape2)
library(DT)
library(readxl)
library(sf)
library(mapview)
library(plotly)

# Bring in custom plotting functions
source('reportFunctions_simple.R')

# Only report 2 decimal places if you take the comment out
#options(digits=2)
```

```{r readInData, echo=FALSE, message=FALSE, warning=FALSE}

# What is the name of the stream?
prettyStreamName <- 'Catawba Creek' ## User Input


# Read in datasets, note these are relative paths vs absolute paths

# field and environmental data
envData <- read_csv("data/environmentalResults_1.csv", col_types = cols(CollectionDateTime = "c")) %>%
  mutate(date= as.Date(as.character(CollectionDateTime),format = '%m/%d/%Y'),
         prettyDate=format(date, format="%B %Y"))

# macro data, only keep Replicate 1 for reporting purposes
bugData <- read_excel("data/benthicResults_1.xlsx") %>%
  dplyr::rename(FamSCI = `Fam SCI`,SampleSeason= `Sample Season`) %>% 
  filter(RepNum == "1")

# Adjust bug season field to factor for plotting purposes
bugData$SampleSeason <- factor(bugData$SampleSeason, levels=unique(as.character(bugData$SampleSeason)) )


# habitat data
habData<- read_excel("data/habitatResults_1.xlsx")
habData$CollDate <- as.Date(as.character(habData$CollDate),format="%Y-%m-%d") # fix date variable

```



## Site Overview

This is the section where you could put some basic description information about the site. You can also start to use the inline text feature and have a generic sentence that is automatically updated each time you enter a new dataset. 

For instance, `r prettyStreamName` was sampled `r nrow(bugData)` times for benthic macroinvertebrates and `r nrow(envData)` times for ambient parameters between `r format(min(envData$date), format="%B %Y")` and `r format(max(envData$date), format="%B %Y")`.

Alternatively, you could turn that information into a table to save space. Here are a few quick stats from the datasets.

```{r Quick Site overview, echo=FALSE}

# Organize bug data
bugTable <- select(bugData, StationID, StreamName, Location, FamSCI) %>% # keep only parameters of interest
  mutate(`Median VSCI` = format(median(FamSCI), digits = 3), # calculate median
         `Number of Samples` = n()) %>% # and the number of samples
  select(-c(FamSCI)) %>% # remove individual scores for overview
  distinct(`Number of Samples`, .keep_all = TRUE) # keep only 1 row

siteTable <- select(envData, StationID, Longitude, Latitude) %>% # keep only parameters of interest
  distinct(StationID, .keep_all = TRUE) %>% # limit to one row
  left_join(bugTable, by = 'StationID') # join with the bugTable informationk with mapview 

kable(t(siteTable), format = "html")

```


### Map Option
And you can give a little interactive map if you'd like.

```{r basic mapview map, fig.width=4, fig.height=4}
# make the data into a spatial object for plotting on a map
siteTable_sf <- st_as_sf(siteTable, coords = c("Longitude", "Latitude"),  # make spatial layer using these columns
                         remove = T, # Remove these lat/lon columns from the dataframe
                         crs = 4326) # add coordinate reference system, needs to be WQS84 to work with mapview

mapview(siteTable_sf,
        label= siteTable_sf$StationID, # What field do you want to populate when you mouse over site?
        layer.name = 'Sample Site', # Rename layer to something pretty for user, siteTable_sf doesnt really mean much to them
        legend=FALSE) # turn off legend in top right corner
# there are many more mapview options, this is just the bare bones

```


***

## Macroinvertebrate Data

Here is a basic plot of Family level Virginia Stream Condition Index scores for the input dataset. The function the plot is built with can handle more samples if you format your data to match  in input data template. You can also go in and alter the code to change SCI threshold to match your benthic cutoff.

```{r , echo=FALSE,message=FALSE, fig.width=5, fig.height=4, fig.cap=paste("Figure 1: Virginia Stream Condition Index (VSCI) for ", prettyStreamName,". A score of less than or equal to  60 indicates an impaired benthic community. No replicate samples are represented in the dataset.",sep="")}


bugDataForPlot <- select(bugData, SampleSeason, FamSCI)


ggplot(bugDataForPlot, aes(x=SampleSeason, y=FamSCI))+
  geom_bar(stat="identity", width = 0.5, position='dodge')+
  labs(x="Sample Season", y="VSCI") +
  theme(axis.text=element_text(size=14, face="bold"), legend.text=element_text(size=14),
        legend.title =   element_text(size=14, face="bold"),
        axis.title=element_text(size=14, face="bold")) +
  scale_y_continuous(name="VSCI", breaks=seq(0, 100, 10),limits=c(0,100)) +
  geom_hline(yintercept=60, color="red", size=1) +  # This is where you would alter where you want your horizontal cutoff
  theme(axis.text.x=element_text(angle=45,hjust=1))

```




#### Individual Metrics

Depending on your audience, you can present the underlying metrics of your SCI and highlight potential metrics driving the overall score. For Virginia, we like to highlight the lowest two metrics. This is a little tricky, but all the JavaScript logic is taken care of for you in the code if you would like to do something similar to your dataset. The overall SCI score is outlined in gray (demonstrating how to change column background color) and logic to highlight scores below a threshold (with bold and red font) is also overviewed in code.


```{r metricScores, echo=FALSE,message=FALSE, fig.width=10, fig.height=4, fig.cap=paste("Table 2: Metric scores used to calculate the VSCI for ", prettyStreamName,". Orange shaded values indicate the two lowest scores that are most likely driving low VSCI scores. Red VSCI scores are below the impairment threshold of 60.",sep="")}

bugScores <- select(bugData, SampleSeason,`%Ephem Score`,`%PT-H Score`,`%Chironomidae Score`,`Fam Richness Score`,
                    `Fam EPT Score`,`Fam %Scraper Score`,`Fam %2Dom Score`,`Fam %MFBI Score`,FamSCI)
                    
bugScores[1:nrow(bugScores),2:10] %>% 
  datatable(rownames =  as.character(bugScores$SampleSeason),
            colnames = c("% Ephem Score","% PT- H Score","% Chironomidae Score","Fam Richness Score",
            "Fam EPT Score","Family %Scraper Score","Family %2 Domniant Score","Family %MFBI Score","VSCI"),
            options=list(dom= 't',
                         rowCallback = # the data.length-1) part makes it only calculate each column but VSCI
                           JS('function(row, data) {
                               var num_data = data.slice(1,data.length-1)
                               num_data.sort(function (a, b) {  return a - b;  });
                               for(i=1;i < data.length; i++) {
                               if(data[i]==num_data[num_data.length-1]) {
                               $("td:eq("+i+")", row).css("background-color", "none")
                               } else if(data[i]==num_data[0]) {
                               $("td:eq("+i+")", row).css("background-color", "orange")
                               } else if(data[i]==num_data[1]) {
                               $("td:eq("+i+")", row).css("background-color", "orange")
                               }}}') # that is the money JavaScript part that highlights the two lowest cells in a row orange
                         )) %>% 
  
  formatStyle("FamSCI", backgroundColor = "lightgray", color=styleInterval(c(60),c('red','black')),
              fontWeight = styleInterval(c(60),c('bold','normal'))) %>% # background color, font color, and font weight alterations
  
  formatRound(c("%Ephem Score","%PT-H Score","%Chironomidae Score","Fam Richness Score","Fam EPT Score","Fam %Scraper Score",
                "Fam %2Dom Score","Fam %MFBI Score","FamSCI"), 2) # round all metrics to two decimal places


```



## Ambient Data

#### pH

You can include some basic plots of all the ambient data you have collected throughout the input dataset. You can also use inline text again to overview the data automatically. E.g. `r prettyStreamName` was sampled for pH data between `r format(min(envData$date),"%Y")` and `r format(max(envData$date),"%Y")`. pH measurements in `r prettyStreamName` ranged from `r min(envData$'pH (unitless)', na.rm=TRUE)` -  `r max(envData$'pH (unitless)', na.rm=TRUE)` with a median value of `r median(envData$'pH (unitless)', na.rm=TRUE)` (n=`r nrow(envData[!is.na(envData$'pH (unitless)'), ])`) (Figure 2). 

Code for a basic ggplot is included. If you have cutoffs for certain parameters, you can go in to the parameter settings section to play with the background colors and zones. Note the difference between the pH plot and the DO plot. The background thresholds add a lot of context if you can add them.

```{r pH, echo=FALSE,message=FALSE, fig.width=10, fig.height=4, fig.cap=paste("Figure 2: Describe pH data here.")}
# basic 
gMasterPlot(envData,'pH (unitless)','3 month')
```

#### Dissolved Oxygen

Dissolved Oxygen in `r prettyStreamName` ranged from `r min(envData$'DO (mg/L)', na.rm=TRUE)` -  `r max(envData$'DO (mg/L)', na.rm=TRUE)` mg/L.  The median dissolved oxygen value was `r median(envData$'DO (mg/L)', na.rm=TRUE)` mg/L (n=`r nrow(envData[!is.na(envData$'DO (mg/L)'), ])`) (Figure 3). 


```{r DO, echo=FALSE,message=FALSE, fig.width=7, fig.height=6, fig.cap=paste("Figure 3: Describe DO data here.")}
# with thresholds
gMasterPlot(envData,'DO (mg/L)','3 month')
```

## Sneaky Way to Save Space {#buttons .tabset .tabset-fade .tabset-pills}

***
Save space in your document using buttons or tabs for sub sections **IF YOU PLAN ON DISTRIBUTING REPORT AS HTML DOCUMENT**. Result can look like this:

### Total Nitrogen 

Total nitrogen in `r prettyStreamName` ranged from `r min(envData$'TN (mg/L)', na.rm=TRUE)` -  `r max(envData$'TN (mg/L)', na.rm=TRUE)` mg/L with a median of `r median(envData$'TN (mg/L)', na.rm=TRUE)` mg/L (n= `r nrow(envData[!is.na(envData$'TN (mg/L)'), ])`) (Figure 4). 

```{r TN, echo=FALSE,message=FALSE, fig.width=9, fig.height=6, fig.cap=paste("Figure 4: Describe TN data here.")}
gMasterPlot(envData,'TN (mg/L)','3 month')
```



***

### Total Dissolved Solids

Total dissolved solids measurements in `r prettyStreamName` ranged from `r min(envData$"TDS (mg/L)", na.rm=TRUE)` -  `r max(envData$"TDS (mg/L)", na.rm=TRUE)` mg/L (Figure 5). The median TDS value was `r median(envData$"TDS (mg/L)", na.rm=TRUE)` mg/L  (n=`r nrow(envData[!is.na(envData$"TDS (mg/L)"), ])`). 

```{r TDS, echo=FALSE,message=FALSE, fig.width=9, fig.height=6, fig.cap=paste("Figure 5: Describe TDS data here.")}
gMasterPlot(envData,"TDS (mg/L)",'3 month')
```

***


### Specific Conductivity

Specific Conductivity measurements in `r prettyStreamName` ranged from `r min(envData$"SpCond (uS/cm)", na.rm=TRUE)` -  `r max(envData$"SpCond (uS/cm)", na.rm=TRUE)` uS/cm (Figure 6). The median specific conductivity value was `r median(envData$"SpCond (uS/cm)", na.rm=TRUE)` uS/cm  (n=`r nrow(envData[!is.na(envData$"SpCond (uS/cm)"), ])`). 

```{r SpCond, echo=FALSE,message=FALSE, fig.width=9, fig.height=6, fig.cap=paste("Figure 6: Describe SpCond data here.")}
gMasterPlot(envData,"SpCond (uS/cm)",'3 month')
```




***

<br><br>

## Total Habitat


If you are planning on distributing report as an HTML document, I recommend using interactive plots with the plotly package. The plots stay interactive even as you distribute the .html file. I find if very useful for larger datasets where users may want to zoom in/out of certain areas and especially with time series data. Below is an example with the total habitat scores. 


`r prettyStreamName` total habitat scores ranged from `r min(envData$'Total Habitat (unitless)', na.rm=TRUE)` -  `r max(envData$'Total Habitat (unitless)', na.rm=TRUE)` and the median value was `r median(envData$'Total Habitat (unitless)', na.rm=TRUE)`(n=`r nrow(envData[!is.na(envData$'Total Habitat (unitless)'), ])`) (Figure 7). 

```{r , echo=FALSE,message=FALSE, fig.width=9, fig.height=6, fig.cap=paste("Figure 7: Describe Total Habitat data here.")}
 
totalHabitat <- filter(envData, !is.na(`Total Habitat (unitless)`)) # filter to just records with total habitat data
totalHabitat$SampleDate <- as.POSIXct(totalHabitat$CollectionDateTime, format="%m/%d/%Y") # convert date time to useable format for plotly
    
box1 <- data.frame(SampleDate = c(min(totalHabitat$SampleDate), min(totalHabitat$SampleDate), max(totalHabitat$SampleDate),max(totalHabitat$SampleDate)), 
                   y = c(150, 250, 250, 150))
box2 <- data.frame(x = c(min(totalHabitat$SampleDate), min(totalHabitat$SampleDate), max(totalHabitat$SampleDate),max(totalHabitat$SampleDate)),
                   y = c(130, 150, 150, 130))
box3 <- data.frame(x = c(min(totalHabitat$SampleDate), min(totalHabitat$SampleDate), max(totalHabitat$SampleDate),max(totalHabitat$SampleDate)), 
                   y = c(100, 130, 130, 100))
box4 <- data.frame(x = c(min(totalHabitat$SampleDate), min(totalHabitat$SampleDate), max(totalHabitat$SampleDate),max(totalHabitat$SampleDate)), 
                   y = c(0, 100, 100, 0))
    
# plotly plot    
plot_ly(data=totalHabitat) %>%
  add_polygons(x = ~SampleDate, y = ~y, data = box1, fillcolor = "firebrick",opacity=0.6, line = list(width = 0),
                   hoverinfo="text", name =paste('High Probability of Stress to Aquatic Life')) %>%
      add_polygons(data = box2, x = ~x, y = ~y, fillcolor = "#F0E442",opacity=0.6, line = list(width = 0),
                   hoverinfo="text", name =paste('Medium Probability of Stress to Aquatic Life')) %>%
      add_polygons(data = box3, x = ~x, y = ~y, fillcolor = "#009E73",opacity=0.6, line = list(width = 0),
                   hoverinfo="text", name =paste('Low Probability of Stress to Aquatic Life')) %>%
      add_polygons(data = box4, x = ~x, y = ~y, fillcolor = "#0072B2",opacity=0.6, line = list(width = 0),
                   hoverinfo="text", name =paste('No Probability of Stress to Aquatic Life')) %>%
      add_markers(data=totalHabitat, x= ~SampleDate, y= ~`Total Habitat (unitless)`,mode = 'scatter',
                  name="Total Habitat (unitless)",marker = list(color= '#535559'),
                  hoverinfo="text",text=~paste(sep="<br>",
                                               paste("Date: ",SampleDate),
                                               paste("Total Habitat : ",`Total Habitat (unitless)`,"(unitless)")))%>%
      layout(showlegend=FALSE,
             yaxis=list(title="Total Habitat (unitless)"),
             xaxis=list(title="Sample Date",tickfont = list(size = 10)))

```


### Total Habitat Heatmap

Further analysis of the the individual metrics (scored 1-20) that make up the qualitative total habitat score are visible in the heatmap below (Table 6). Darker colors indicate lower habitat metric scores.

```{r habitatHeatmap, echo=FALSE,message=FALSE, fig.cap="Table 2: Qualitative habitat measurements taken during biological monitoring. The lighter red colors represent habitat scores that are optimal or suboptimal and progressively darker red colors represent marginal or poor habitat scores."}

# Set how many colors you will use and call them out by hex name
brks <- 1:19
clrs <- c("#8B0000", "#9D0000", "#AF0000", "#C10000", "#D40000", "#E60000", "#F80000", "#FF1415", "#FF3235", "#FF5055", "#FF6F75",
          "#FF8D95", "#FFABB5", "#FFC3CD", "#FFCDD5", "#FFD7DE", "#FFE1E6", "#FFEBEE", "#FFF5F6", "#FFFFFF")


totHabData <- select(habData,StationID,CollDate,TotHabSc)[!duplicated(select(habData,StationID,CollDate,TotHabSc)),] # get unique sample dates

habDataForTable <- select(habData,-c(TotHabSc)) %>% # these two fields can cause multiple rows for each date in the spread() step, so get rid of them for now
  filter(HabParameter %in% c('ALTER','BANKS','BANKVEG','COVER','EMBED','FLOW','RIFFLES','RIPVEG','SEDIMENT',
                             'SUBSTRATE','VELOCITY')) %>% # get only high gradient parameters 
  spread(HabParameter,HabValue) %>%
  filter(!is.na(EMBED) & !is.na(RIFFLES) & !is.na(VELOCITY)) %>% # if both methods used then get rid of sample dates not sampled as high gradient
  rowwise()%>%
  left_join(totHabData, by=c("StationID","CollDate")) %>%
  dplyr::rename(TotalHabitat = TotHabSc) %>%
  arrange(CollDate)

DT::datatable(habDataForTable,escape=F, rownames = F, colnames = c('Station ID','Date','Channel Alteration','Banks', 
                                                                   'Bank Vegetation', 'Embeddedness', 
                                                                   'Flow', 'Riffles', 'Riparian Vegetation', 
                                                                   'Sediment', 'Substrate','Velocity', 'Total Habitat'),
              options=list(pageLength=nrow(habDataForTable),dom= 'Bt' )) %>%
  formatStyle(names(habDataForTable)[3:12], backgroundColor = styleInterval(brks, clrs), 
              textAlign = 'center', `font-family` = 'Arial') %>%
  formatStyle(names(habDataForTable), fontWeight = styleInterval(10, c('bold','normal')), 
              textAlign = 'center', `font-family` = 'Arial') %>%
  formatStyle('TotalHabitat', backgroundColor = "lightgray")


```
